---
title: "p8105_hw2_xs2529"
author: Xinyi Shang
output: html_document
date: "2023-10-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load_libraries}
library(tidyverse)
library(readxl)
```

## Problem 1

```{r clean_538_pols}
month_df = 
  tibble(
    month_num = 1:12,
    month_abb = month.abb,
    month = month.name
  )

pols = 
  read_csv("./data/fivethirtyeight_datasets/pols-month.csv") |>
  separate(mon, into = c("year", "month_num", "day"), convert = TRUE) |>
  mutate(
    president = recode(prez_gop, "0" = "dem", "1" = "gop", "2" = "gop")) |>
  left_join(x = _, y = month_df) |> 
  select(year, month, everything(), -day, -starts_with("prez")) 

```

We also clean the 538 `snp` data, which contains information related to Standard & Poor’s stock market index.

```{r clean_538_snp}

snp = 
  read_csv("./data/fivethirtyeight_datasets/snp.csv", col_types = cols(date = col_date(format = "%m/%d/%y"), close = col_double())) |> #modified for correct date
  separate(date, into = c("year", "month", "day"), convert = TRUE) |> #modified for correct date
  mutate(year = ifelse(year > 2023, year - 100,year)) |> #modified for correct date
  arrange(year, month) |>
  mutate(month = month.name[month]) |>
  select(year, month, close) 

```

Finally, we tidy the `unemployment` data so that it can be merged with the `pols` and `snp` datasets.

```{r clean_538_unemp}
unemployment = 
  read_csv("./data/fivethirtyeight_datasets/unemployment.csv") |>
  rename(year = Year) |>
  pivot_longer(
    Jan:Dec, 
    names_to = "month_abb",
    values_to = "unemployment"
  ) |> 
  left_join(x = _, y = month_df) |> 
  select(year, month, unemployment)
```

Now we merge the three datasets!

```{r merge_538}
data_538 = 
  left_join(pols, snp) |>
  left_join(x = _, y = unemployment)

str(data_538)

write.csv(data_538, "data_538.csv", row.names=FALSE) #save the data into a csv file
```

## Problem 2

```{r clean Mr TrashWheel}
Mr_TrashWheel = 
  read_excel("./data/202207 Trash Wheel Collection Data.xlsx",sheet = "Mr. Trash Wheel", col_names = TRUE) |>
  select(c(1:14)) |> #ignore the column without data
  janitor::clean_names() |> #clean the dataset
  drop_na(dumpster) |> #remove data without dumpster number
  rename (volume = volume_cubic_yards) |>
  rename (weight = weight_tons) |>
  mutate (homes_powered, homes_powered = weight*500/30) |> #calculate number of homes powered
  mutate(year = as.numeric(year)) |>
  mutate(TrashWheel = "MrTrashWheel")
```


```{r clean Professor TrashWheel}
Prof_TrashWheel = 
  read_excel("./data/202207 Trash Wheel Collection Data.xlsx",sheet = "Professor Trash Wheel", col_names = TRUE) |>
  janitor::clean_names() |> #clean the dataset
  drop_na(dumpster) |> #remove data without dumpster number
  rename (volume = volume_cubic_yards) |>
  rename (weight = weight_tons) |>
  mutate (homes_powered, homes_powered = weight*500/30) |> #calculate number of homes powered
  mutate(year = as.numeric(year)) |>
  mutate(TrashWheel = "ProfTrashWheel")

```


```{r clean Gwynnda TrashWheel}
Gwynnda_TrashWheel = 
  read_excel("./data/202207 Trash Wheel Collection Data.xlsx",sheet = "Gwynnda Trash Wheel", col_names = TRUE) |>
  janitor::clean_names() |> #clean the dataset
  drop_na(dumpster) |> #remove data without dumpster number
  rename (volume = volume_cubic_yards) |>
  rename (weight = weight_tons) |>
  mutate (homes_powered, homes_powered = weight*500/30) |> #calculate number of homes powered
  mutate(year = as.numeric(year))|>
  mutate(TrashWheel = "GwynndaTrashWheel")

```

```{r merge dataset}
TrashWheel_combined = bind_rows (Mr_TrashWheel,Prof_TrashWheel,Gwynnda_TrashWheel)
```

The total number of observation after cleaning is `r nrow(TrashWheel_combined)`, and the variables of this dataset are `r names(TrashWheel_combined)`. 

The total weight of trash collected by Professor Trash Wheel is `r filter(TrashWheel_combined, TrashWheel == "ProfTrashWheel") |> pull(weight)|> sum()`

The total number of cigarette butts collected by Gwynnda in July of 2021 is `r filter(TrashWheel_combined, TrashWheel == "GwynndaTrashWheel", year == "2021", month == "July") |> pull(cigarette_butts) |> sum() |> as.integer()`.

## Problem 3
```{r clean the baseline}
baseline = 
  read_csv("./data/data_mci/MCI_baseline.csv", skip = 1) |>
  janitor::clean_names() |> #clean the dataset
  mutate (sex = ifelse(sex == 1, "Male", "Female")) |>
  mutate (apoe4 = ifelse(apoe4 == 1, "APOE4 carrier", "APOE4 non-carrier")) |>
  mutate (age_at_onset = as.numeric(age_at_onset))
```

```{r remove participants who are MCI free}
baseline_clean = 
  baseline |>
  filter(age_at_onset > current_age | is.na(age_at_onset))

```

```{r clean the amyloid}
amyloid = 
  read_csv("./data/data_mci/mci_amyloid.csv", skip = 1) |>
  janitor::clean_names() |> #clean the dataset 
  rename(id = study_id)
```

```{r remove participants who have missing values}
amyloid_clean = 
  amyloid |>
  mutate(amyloid = ifelse(is.numeric(amyloid == FALSE),NA,amyloid)) |>
  drop_na(amyloid)  #remove participants with any missing visit
```


```{r combine dataset}
combined_full = 
  full_join(baseline,amyloid, by = "id") #total participants in this study 

combined_clean_full = 
  full_join(baseline_clean,amyloid_clean) #participants without missing value in either dataset

combined_clean = 
  inner_join(baseline_clean,amyloid_clean) #participants without missing value

```

In total, there are `r nrow(combined_full)` participants who attended this study. There are `r nrow(combined_full)-nrow(baseline)` participants only appears in the amyloid dataset, and `r nrow(combined_full)-nrow(amyloid)` participants only appears in the baseline dataset.

After cleaning the data, there are `r nrow(combined_clean_full)` participants who attended this study without missing value in either dataset. There are `r nrow(combined_clean_full)-nrow(baseline_clean)` participants only appears in the amyloid dataset, and `r nrow(combined_clean_full)-nrow(amyloid_clean)` participants only appears in the baseline dataset.

After cleaning, there are `r nrow(combined_clean)` participatns appears in both dataset. Among all the participants, `r combined_clean |> pull(`

Q1: Cleaning: what data should be removed for baseline and amyloid?
baseline: having MCI at baseline (should i remove participants who never develop MCI?)
amyloid: removed participatns have NA values (missing visit)


Check whether some participants appear in only the baseline or amyloid datasets, and comment on your findings.  (do this with cleaned dataset or the raw one?)
Combine the demographic and biomarker datasets so that only participants who appear in both datasets are retained, and briefly describe the resulting dataset; export the result as a CSV to your data directory.
