---
title: "p8105_hw2_xs2529"
author: Xinyi Shang
output: html_document
date: "2023-10-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,collapse = TRUE, message = FALSE)
```

```{r load_libraries}
library(tidyverse)
library(readxl)
```

## Problem 1

```{r clean_538_pols}
month_df = 
  tibble(
    month_num = 1:12,
    month_abb = month.abb,
    month = month.name
  )

pols = 
  read_csv("./data/fivethirtyeight_datasets/pols-month.csv") |>
  separate(mon, into = c("year", "month_num", "day"), convert = TRUE) |>
  mutate(
    president = recode(prez_gop, "0" = "dem", "1" = "gop", "2" = "gop")) |>
  left_join(x = _, y = month_df) |> 
  select(year, month, everything(), -day, -starts_with("prez")) 

```

We also clean the 538 `snp` data, which contains information related to Standard & Poor’s stock market index.

```{r clean_538_snp}

snp = 
  read_csv("./data/fivethirtyeight_datasets/snp.csv", col_types = cols(date = col_date(format = "%m/%d/%y"), close = col_double())) |> #modified for correct date
  separate(date, into = c("year", "month", "day"), convert = TRUE) |> #modified for correct date
  mutate(year = ifelse(year > 2023, year - 100,year)) |> #modified for correct date
  arrange(year, month) |>
  mutate(month = month.name[month]) |>
  select(year, month, close) 

```

Finally, we tidy the `unemployment` data so that it can be merged with the `pols` and `snp` datasets.

```{r clean_538_unemp}
unemployment = 
  read_csv("./data/fivethirtyeight_datasets/unemployment.csv") |>
  rename(year = Year) |>
  pivot_longer(
    Jan:Dec, 
    names_to = "month_abb",
    values_to = "unemployment"
  ) |> 
  left_join(x = _, y = month_df) |> 
  select(year, month, unemployment)
```

Now we merge the three datasets!

```{r merge_538}
data_538 = 
  left_join(pols, snp) |>
  left_join(x = _, y = unemployment)

str(data_538)

write.csv(data_538, "data_538.csv", row.names=FALSE) #save the data into a csv file
```

Notice that there are some `NA` values in the `close` and `unemployment` variables, which indicate that the value of these variables is missing at those locations.

Let's talk about the 538 datasets. The `pols` data has `r nrow(pols)` observations and `r ncol(pols)` variables and tells us about the party affiliation distribution (democrat or republican) for governors and senators for a given year from years `r (pols |> select(year) |> range())[1]` to `r (pols |> select(year) |> range())[2]`. It also tells us whether the sitting president was a democrat or republican. The `snp` data has `r nrow(snp)` observations and `r ncol(snp)` variables, ranging from years `r (snp |> select(year)|> range())[1]` to `r (snp |> select(year) |> range())[2]`. The `unemployment` data has `r nrow(unemployment)` observations and `r ncol(unemployment)` variables ranging from years `r (unemployment |> select(year) |> range())[1]` to `r (unemployment |> select(year) |> range())[2]`. In Januarys in or after 1975 in which a democrat was president, the average unemployment rate was `r filter(data_538, month == "January", year >= 1975, president == "dem") |> pull(unemployment) |> mean() |> round(2)`.  The average unemployment rate over the same time period in which a republican was president was `r filter(data_538, month == "January", year >= 1975, president == "gop") |> pull(unemployment) |> mean() |> round(2)`.


## Problem 2

Firstly, I cleaned the Mr. TrashWheel, Professor TrashWheel, and Gwynnda TrashWheel dataset. 
```{r clean Mr TrashWheel}
Mr_TrashWheel = 
  read_excel("./data/202309 Trash Wheel Collection Data.xlsx",sheet = "Mr. Trash Wheel", col_names = TRUE) |>
  select(c(1:14)) |> #ignore the column without data
  janitor::clean_names() |> #clean the dataset
  drop_na(dumpster) |> #remove data without dumpster number
  rename (volume = volume_cubic_yards) |>
  rename (weight = weight_tons) |>
  mutate (homes_powered, homes_powered = weight*500/30) |> #calculate number of homes powered
  mutate(year = as.numeric(year)) |>
  mutate(TrashWheel = "MrTrashWheel")
```


```{r clean Professor TrashWheel}
Prof_TrashWheel = 
  read_excel("./data/202309 Trash Wheel Collection Data.xlsx",sheet = "Professor Trash Wheel", col_names = TRUE) |>
  janitor::clean_names() |> #clean the dataset
  drop_na(dumpster) |> #remove data without dumpster number
  rename (volume = volume_cubic_yards) |>
  rename (weight = weight_tons) |>
  mutate (homes_powered, homes_powered = weight*500/30) |> #calculate number of homes powered
  mutate(year = as.numeric(year)) |>
  mutate(TrashWheel = "ProfTrashWheel")

```


```{r clean Gwynnda TrashWheel}
Gwynnda_TrashWheel = 
  read_excel("./data/202309 Trash Wheel Collection Data.xlsx",sheet = "Gwynnda Trash Wheel", col_names = TRUE) |>
  janitor::clean_names() |> #clean the dataset
  drop_na(dumpster) |> #remove data without dumpster number
  rename (volume = volume_cubic_yards) |>
  rename (weight = weight_tons) |>
  mutate (homes_powered, homes_powered = weight*500/30) |> #calculate number of homes powered
  mutate(year = as.numeric(year))|>
  mutate(TrashWheel = "GwynndaTrashWheel")

```

Then, I merged those three datasets together. 
```{r merge dataset}
TrashWheel_combined = bind_rows (Mr_TrashWheel,Prof_TrashWheel,Gwynnda_TrashWheel)

str(TrashWheel_combined)
```

The total number of observation after cleaning is `r nrow(TrashWheel_combined)`, and the variables of this dataset are `r names(TrashWheel_combined)`. 

The dataset contains the trashes removed by the water-wheel vessel from the Inner Harbor in Baltimore from `r (TrashWheel_combined |> select(year) |> range())[1]` to `r (TrashWheel_combined |> select(year) |> range())[2]`. Three types of Trash Wheel are included in this dataset: Mr. Trash Wheel, Professor Trash Wheel, and Gwynnda Trash Wheel. The total weight of trashes collected by those three trash wheels is `r TrashWheel_combined |> pull(weight)|> sum()` tons.

The total weight of trash collected by Professor Trash Wheel is `r filter(TrashWheel_combined, TrashWheel == "ProfTrashWheel") |> pull(weight)|> sum()` tons. 

The total number of cigarette butts collected by Gwynnda in July of 2021 is `r filter(TrashWheel_combined, TrashWheel == "GwynndaTrashWheel", year == "2021", month == "July") |> pull(cigarette_butts) |> sum() |> as.integer()`. 

## Problem 3

#### Baseline dataset
Firstly, I cleaned the baseline dataset.
```{r clean the baseline}
baseline = 
  read_csv("./data/data_mci/MCI_baseline.csv", skip = 1) |>
  janitor::clean_names() |> #clean the dataset
  mutate (sex = ifelse(sex == 1, "Male", "Female")) |>
  mutate (apoe4 = ifelse(apoe4 == 1, "APOE4 carrier", "APOE4 non-carrier")) |>
  mutate (age_at_onset = as.numeric(age_at_onset))
```

Then, I removed the participants who did not meet the inclusion criteria, which is no MCI at the baseline. 
```{r remove participants who are MCI free}
baseline_clean = 
  baseline |>
  filter(age_at_onset > current_age | is.na(age_at_onset))
#only participants who develop MCI after the baseline and participants who never develop MCI are kept. 
```

In the baseline dataset, there are `r nrow(baseline)` participants recruited in this study, and `r nrow(baseline_clean)` participants meet the inclusion criteria. 

Among all the participants who meet the criteria, there are `r baseline_clean |> filter(is.na(age_at_onset) == FALSE) |> nrow()` participants who develope MCI. The average baseline age is `r baseline_clean |> pull(current_age) |> mean() |>round(2)`.

There are `r (nrow(filter(baseline, sex == "Female", apoe4 == "APOE4 carrier")) / nrow(filter(baseline, sex == "Female")) *100) |> round(2)`% of women in the study are APOE4 carriers.

#### Amyloid dataset
Firstly, I cleaned the amyloid dataset. 
```{r clean the amyloid}
amyloid_clean = 
  read_csv("./data/data_mci/mci_amyloid.csv", skip = 1) |>
  janitor::clean_names() |> #clean the dataset
  rename(id = study_id)
```

Then, I performed pivot for the amyloid dataset. 
```{r perform pivot for the dataset}
amyloid_pivot = 
  pivot_longer(amyloid_clean, col = !id, names_to = "time", values_to = "amyloid beta 42/40 ratio")
  
```

There are `NA` values in the `amyloid beta 42/40 ratio`, which means there are missing values.

In total, there are `r nrow(amyloid_clean)` participants recruited in this study, and `r nrow(amyloid_clean)` participants meet the inclusion criteria for the amyloid dataset. The variables evaluated are `r names (amyloid_clean)`.

#### Merge dataset
In the end, I combined the baseline and the amyloid dataset. 
```{r combine dataset}
combined_full = 
  full_join(baseline,amyloid_clean, by = "id") #total participants in this study 

combined_clean_full = 
  full_join(baseline_clean,amyloid_clean, by = "id") #participants meet all criteria

combined_clean = 
  inner_join(baseline_clean,amyloid_clean, by = "id") #participants meet all criteria and appears in both dataset

combined_pivot = 
  inner_join(baseline_clean,amyloid_pivot, by = "id") #combine the dataset after pivot

str(combined_pivot)

```

After cleaning the data, there are `r nrow(combined_clean_full)` participants who attended this study who fullfilled the inclusion criteria. There are `r anti_join(amyloid_clean, baseline_clean, by = "id" ) |> nrow()` participants only appears in the amyloid dataset, and `r anti_join(baseline_clean, amyloid_clean, by = "id") |> nrow()` participants only appears in the baseline dataset.

After pivoting the amyloid dataset, the final dataset have `r nrow(combined_pivot)` rows and `r ncol(combined_pivot)` columns.The variables considered in this dataset are `r names(combined_pivot)`.

After cleaning, there are `r nrow(combined_clean)` participatns appears in both dataset and satisfied all inclusion criteria. 
Among those participants, there are `r filter(combined_clean, sex == "Female") |> nrow()` women and `r filter(combined_clean, sex == "Male") |> nrow()` men. The average age at baseline is `r combined_clean |> pull(current_age) |> mean() |> round(2)`. 